| Base de datos de abejas Ibéricas \n| Iberian Bees database\n

### 2022-03-16 Manuscript/manuscript.Rmd
Fixing typo

Here we compile an Iberian Bee Database of occurrence records by integrating digitalization efforts by leading national history museums, with more than 100 individual datasets contributed by researchers, and publicly available information on online repositories. After data cleaning and harmonization, we release the first `r nrow(data)` records from the Iberian Peninsula and the Balearic Islands. Figure 1 shows the spatial distribution of records, which as expected its biased to certain areas near main cities (Fig 1). Overall, the database contains information of `r length(unique(data$Accepted_name))` different species, for which the most commonly collected are depictted in Figure 2. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting this taxa (Figure 3A). Figure 3 also shows the temporal distribution of records within years (Figure 3B). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where it was collected. This Iberian Bee Database (v.1.0) will continue to grow, but it is the starting point to better understand and conserve bee biodiversity. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

Con la ausencia de esfuerzos de muestreo estandarizados se podría pensar que información detallada de la ocurrencia de especies en el espacio y en el tiempo no existe. Sin embargo, existen diversas fuentes con información muy útil. En primer lugar, los museos de historia natural albergan colecciones históricas de naturalistas amateur e investigadores que coleccionaron abejas [@bartomeus2019]. Puesto que las colecciones de museos suelen estar muy bien clasificadas y etiquetadas, es posible recopilar la información sobre los especímenes de la colección y cuándo y dónde estos fueron capturados. En segundo lugar, la Península Ibérica tiene una rica tradición de investigadores sobre polinizadores y polinización [@archer2014] que han recopilado gran cantidad información de abejas con diferentes objetivos. Finalmente, muchos datos de ocurrencias han sido liberados de forma abierta a través de portales de internet que centralizan datos almacenados por instituciones de todo el mundo (p.ej. Gbif) o datos de ciencia ciudadana (p.ej.iNaturalist). No obstante, sigue siendo un desafio digitalizar, acceder y limpiar toda la información existente de las diferentes fuentes de datos.

>>>>

Here we compile an Iberian Bee Database of occurrence records by integrating digitalization efforts by leading national history museums, with more than 100 individual datasets contributed by researchers, and publicly available information on online repositories. After data cleaning and harmonization, we release the first `r nrow(data)` records from the Iberian Peninsula and the Balearic Islands. Figure 1 shows the spatial distribution of records, which as expected its biased to certain areas near main cities (Figure 1). Overall, the database contains information of `r length(unique(data$Accepted_name))` different species, for which the most commonly collected are depicted in Figure 2. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting this taxa (Figure 3A). Figure 3 also shows the temporal distribution of records within years (Figure 3B). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where it was collected. This Iberian Bee Database (v.1.0) will continue to grow, but it is the starting point to better understand and conserve bee biodiversity. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

Con la ausencia de esfuerzos de muestreo estandarizados se podría pensar que información detallada de la ocurrencia de especies en el espacio y en el tiempo no existe. Sin embargo, existen diversas fuentes con información muy útil. En primer lugar, los museos de historia natural albergan colecciones históricas de naturalistas amateur e investigadores que coleccionaron abejas [@bartomeus2019]. Puesto que las colecciones de museos suelen estar muy bien clasificadas y etiquetadas, es posible recopilar la información sobre los especímenes de la colección y cuándo y dónde estos fueron capturados. En segundo lugar, la Península Ibérica tiene una rica tradición de investigadores sobre polinizadores y polinización [@archer2014] que han recopilado gran cantidad información de abejas con diferentes objetivos. Finalmente, muchos datos de ocurrencias han sido liberados de forma abierta a través de portales de internet que centralizan datos almacenados por instituciones de todo el mundo (p.ej. Gbif) o datos de ciencia ciudadana (p.ej. iNaturalist). No obstante, sigue siendo un desafio digitalizar, acceder y limpiar toda la información existente de las diferentes fuentes de datos.
